---
title: "CEFI"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[NOAA's Physical Science Laboratory (PSL)](https://psl.noaa.gov/) [Climate Ecosystems and Fisheries Initiative Portal](https://psl.noaa.gov/cefi_portal/) serves historical and forecast data useful in ecological studies. Data is served using a [THREDDS](https://psl.noaa.gov/thredds/catalog/Projects/CEFI/regional_mom6/catalog.html) catalog, but PSL also makes [tabular catalogs](https://psl.noaa.gov/cefi_portal/var_list_northwest_atlantic_hist_run.html) available, too.

# Requirements

 + [R v4.1+](https://www.r-project.org/)
 + [rlang](https://CRAN.R-project.org/package=rlang)
 + [dplyr](https://CRAN.R-project.org/package=dplyr)
 + [sf](https://CRAN.R-project.org/package=sf)
 + [stars](https://CRAN.R-project.org/package=stars)
 + [jsonlite](https://CRAN.R-project.org/package=jsonlite)
 + [tidync](https://CRAN.R-project.org/package=tidync)

# Installation

```
remotes::install_github("BigelowLab/cefi")
```

# Usage

Load the libraries needed.
```{r}
suppressPackageStartupMessages({
  library(rnaturalearth)
  library(cefi)
  library(tidync)
  library(stars)
  library(dplyr)
})
```

## Catalogs

CEFI offers THREDDS catalogs which are great for mining programmatically, but they also provide simple tabular catalogs: one for historical runs and one for forecasts.  (Yes, the first variable is "Varible_Name".)  These are easy for users to navigate. 

```{r}
uri = catalog_uri(region = "Northwest Atlantic", period = "history")
hist = read_catalog(uri) |>
  dplyr::glimpse()
```
```{r}
uri = catalog_uri(region = "Northwest Atlantic", period = "forecast")
fcst = read_catalog(uri) |>
  dplyr::glimpse()
```

Assuming these catalogs will remain and stay up-to-date, we can leverage them in lieu of coding up software to navigate the THREDDS catalogs.

# Getting data

To get data select one row from either catalog, and open that resource which we see in R as a [tidync](https://docs.ropensci.org/tidync/) object useful for navigating and extracting netcdf files.

```{r}
nc = hist |>
  dplyr::filter(Variable_Name == "btm_o2") |>
  cefi_open()
nc
```

Note that the spatial dimensions of the "active" grid are defined by `xh` nd `yh` not `x` and `y` or `lon` and `lat`.  That is a choice of the creator of the NetCDF resource, and so we use those names for filtering.

Next we filter the array with `cefi_filter()`, so that we can collect a subset of the data. This function is a wrapper around the `tidync::hyper_filter()` function; we wrap because the time coordinate in the NetCDF files is not in user-friendly format.  Note that it is important that `time` comes before any other filtering element.

```{r}
nc = cefi_filter(nc, 
                 time = as.Date(c("1995-12-16", "1995-12-20")), 
                 xh = xh > -75 & xh < -60,
                 yh = yh > 40 & yh < 50)
```

It is important to understand that the actual data is not loaded into R (yet), think of this a prefiltering step.  To actually get the data we can call `cefi_var`.  You can explore more about prefiltering at the [tidync website](https://docs.ropensci.org/tidync/).

Now we can laod the data into R, using `cefi_var()`.

```{r}
s = cefi_var(nc)
s
```

And finally we can plot the result.

```{r}
coast = rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") |>
  sf::st_geometry() |>
  sf::st_crop(sf::st_bbox(s)) 

plot_coast = function(){
  plot(sf::st_geometry(coast), add = TRUE, col = "darkorange")
}

plot(s, hook = plot_coast, key.pos = NULL)
```
