---
title: "stars issues"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Preliminaries...

```{r}
suppressPackageStartupMessages({
  library(rnaturalearth)
  library(cefi)
  library(tidync)
  library(stars)
  library(dplyr)
})
uri = "http://psl.noaa.gov/thredds/dodsC/Projects/CEFI/regional_mom6/northwest_atlantic/hist_run/ocean_cobalt_daily_2d.19930101-20191231.btm_o2.nc"
tnc = tidync::tidync(uri) |>
  tidync::hyper_filter(time = time <= 2)
```

```{r}
coast = rnaturalearth::ne_coastline(scale = "medium", returnclass = 'sf')
plot_coast = function(){
  plot(sf::st_geometry(coast), col = "darkorange", add = TRUE)
}

```

# Extract 

## Using an adaption of [MS method](https://github.com/ropensci/tidync/issues/68#issuecomment-484773118)

This approach digs into unexported functions in both `stars`  and `tidync` packages.
```{r}
s1 = tidync_as_stars(tnc) |>
  sf::st_set_crs(sf::st_crs(coast))
```

## Using the naive approach

This approach uses a simpler (but less robust) approach to building a rectilinear grid.

```{r}
s2 = cefi_var(tnc)
```



```{r}
coast2 = coast
plot(s1, hook = plot_coast)
```

```{r}
plot(s2, hook = plot_coast)
```

So, regardless of method we have something unexpected happening in the St Lawrence River region.

# Examine the y-coordinates

Extract the coordinates and then plot the values, and then the consecutive differences.
```{r}
ax = tidync::hyper_transforms(tnc)
yc = ax[['yh']] |> dplyr::pull(1)
plot(yc)
```

```{r}
plot(diff(yc))
```

Clearly something is going on starting at index 602 (47 degrees north) 